---
title: "US Presidential Election 2012-2016 Analysis"
author: "Sunny Nagdev"
date: "December 12, 2018"
output:
  html_document:
    self_contained: no
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




&nbsp;
&nbsp;
&nbsp;



## Dataset Details

Chosen dataset provides Presidential Election Statistics from election year 2012 and 2016. 
Election statistics are merged together with County level information on population, gender, race, income. employment and education, density, etc.

Total Number of Rows in Dataset: 3112.
Total Number of Columns in Dataset: 81.

&nbsp;
&nbsp;


## Data Preparation


#### Data Source: https://www.kaggle.com/joelwilson/2012-2016-presidential-elections
#### Data File: votes.csv
#### Data Dictionary: county_facts_dictionary.csv




&nbsp;
&nbsp;
&nbsp;


```{r code2, warning=FALSE, message=FALSE,echo=FALSE}

## Reading Source File

ER <- read.csv('/Users/Sunny/Desktop/Voting/votes.csv', header = TRUE)

#### Disregarding Columns which are not required

ER$SBO001207 <- NULL;ER$SBO315207 <- NULL;ER$SBO115207 <- NULL;ER$SBO215207 <- NULL;
ER$SBO515207  <- NULL;ER$SBO415207  <- NULL;ER$SBO015207  <- NULL;
ER$MAN450207 <- NULL;ER$WTN220207 <- NULL;ER$RTN130207 <- NULL;
ER$RTN131207 <- NULL;ER$AFN120207 <- NULL;ER$BPS030214 <- NULL;
ER$X.1 <- NULL;ER$X <- NULL;ER$Clinton_Obama <- NULL;ER$Trump_Romney <- NULL;
ER$Trump_Prediction <- NULL; ER$Clinton_Prediction <- NULL; ER$Trump_Deviation <- NULL;
ER$Clinton_Deviation <- NULL



ER1 <- data.frame(ER_county_ID = ER$combined_fips,
                  ER_county_nm = ER$county_name,
                  ER_states = ER$state_abbr,
                  ER_votes_Dem16 = ER$votes_dem_2016,
                  ER_votes_GOP16 = ER$votes_gop_2016,
                  ER_votes_total =  ER$total_votes_2016,
                  ER_Clinton  = ER$Clinton * 100,
                  ER_Trump = ER$Trump * 100,
                  ER_diff2016 = 100 * ER$per_point_diff_2016,
                  ER_diff2012 = 100 *  ER$per_point_diff_2012,
                  ER_Obama = ER$Obama * 100,
                  ER_Romney = ER$Romney * 100,
                  ER_votes_Dem12 = ER$votes_dem_2012,
                  ER_votes_GOP12 = ER$votes_gop_2012,
                  ER_votes_total = ER$total_votes_2012,
                  ER_Pop14 = ER$population2014,
                  ER_Pop10 = ER$population2010,
                  ER_Age_U5_14 = ER$AGE135214,
                  ER_Age_U18_14 = ER$AGE295214,
                  ER_Age_O65_14 = ER$age65plus,
                  ER_Vot_Turnarnd = (ER$total_votes_2016/((100 - (ER$AGE135214 + ER$AGE295214))                                     * ER$population2014/100)) * 100,
                  ER_Gender = ER$SEX255214,
                  ER_Ethn_White = round(ER$White * ER$population2014),
                  ER_Ethn_Black = round(ER$Black * ER$population2014),
                  ER_Amer_Ind = round(ER$RHI325214/100 * ER$population2014),
                  ER_Asian = round(ER$RHI425214/100 * ER$population2014),
                  ER_Hawai = round(ER$RHI525214/100 * ER$population2014),
                  ER_Misc = round(ER$RHI625214/100 * ER$population2014),
                  ER_Hispanic = round(ER$Hispanic/100 * ER$population2014),
                  ER_College_Pop = (ER$Edu_batchelors/100) * ER$population2014,
                  ER_House_income = ER$INC110213,
                  ER_Density = ER$Density)

```

&nbsp;
&nbsp;


## Analysis Objective

1. Univariate Analysis
&nbsp;

       - Determine States where counties got flipped in 2016 Elections.
       - Identify trends for States with significant county flips in 2016 Presidential elections.
       - Examine Ethnic Split and College Populationfor above states.
       - Determine Voter Turnaround across 50 US states.
       - Summarize Presidential Candidates (from 2012 and 2016 elections) Vote % across all US 
         counties.
&nbsp;

2. Bivariate Multivariate Analysis
&nbsp;

  - Examine Voting trends and analyze correlation between vote % for below variables for 2016 and 2012 at county level:
       - College Population 
       - Household Income
       - Female Population
       - Ethnicity
      
3. Analyzing some key distribution
&nbsp;

      - Examine Age Distribution in each state.
      - Point Differenential between 2016 and 2012 presidential elections.
&nbsp;
      
4. Central Limit Theorem Application
&nbsp;

      - Getting Distribution of Income across US States.
      - Drawing 100 Samples for various sample sizes of 10,20,30,40
&nbsp;

5. Various Sampling Techniques
&nbsp;

   - Compare Original Voter Turnaround Percentage across US counties against below Sampling      
        methods.

      - Simple Random Sampling without Replacement
      - Systematic Sampling
      - Stratified Sampling
    
    
&nbsp;
&nbsp;
&nbsp;



## Univariate Analysis 

&nbsp;
&nbsp;


### Analyzing Flipped States in 2016 presidential Elections
    
```{r code45, warning=FALSE, message=FALSE,echo=FALSE}
library(plotly)
library(ggplot2)
library(UsingR)
library(prob)
library(dplyr)
library(knitr)
library(sampling)

### Analyzing Categorical Variable ####################################################

##Identifying Winner for each county in 2016 Presidential Election
ER1$Winner2016 <- ifelse(ER1$ER_votes_Dem16>ER1$ER_votes_GOP16 ,"Clinton","Trump")

##Identifying Winner for each county in 2012 Presidential Election
ER1$Winner2012 <- ifelse(ER1$ER_votes_Dem12>ER1$ER_votes_GOP12 ,"Obama","Romney")


##Identifying Counties flipped by Trump and Clinton in 2016 election

ER1$RepubFlip <- ifelse(ER1$Winner2016 == 'Trump' & ER1$Winner2012 == 'Obama', 1, 0)
ER1$DemoFlip <- ifelse(ER1$Winner2016 == 'Clinton' & ER1$Winner2012 == 'Romney', 1, 0)


##

county_flipR <- subset(ER1,ER1$Winner2016 == 'Trump') 
county_flipR <- subset(county_flipR,county_flipR$Winner2012 == 'Obama')

## Getting subsets for Democratic Winner
county_Demo_Winner <- subset(ER1,ER1$Winner2016 == 'Clinton')

## Getting subsets for Republican Winner
county_Rep_Winner <- subset(ER1,ER1$Winner2016 == 'Trump')


##Identifying Counties flipped by Clinton in 2016 election

county_flipD = subset(ER1,ER1$Winner2016 == 'Clinton')
county_flipD = subset(county_flipD,county_flipD$Winner2012 == 'Romney')


## Configuring the levels for States to show on x axis else all 51 states will be displayed with no bars

county_flipR$ER_states <- factor(county_flipR$ER_states, levels =               c(as.character(unique(county_flipR$ER_states))))

county_flipD$ER_states <- factor(county_flipD$ER_states, levels = c(as.character(unique(county_flipD$ER_states))))


p <- plot_ly(data = county_flipR, 
             x = county_flipR$ER_states, 
             type = "histogram",
             name = 'Flipped By Republicans')%>%
     add_trace(data = county_flipD, 
               x = county_flipD$ER_states, 
               type = "histogram",
               name = 'Flipped by Democrats') %>%
  layout(yaxis = list(title = 'No. of US Counties Flipped in 2016'))%>%
  layout(xaxis = list(title = 'States')); p



cat("Total Number of Counties Flipped by Trump      :", nrow(subset(ER1, ER1$RepubFlip == 1)))



cat("Total Number of Counties Flipped by Clinton    :", nrow(county_flipD))


#######################################################
##Calculating % county flips for Republicans

ER1$LineVal <- 1

county_st <- ER1 %>%
  group_by(ER_states) %>%
  summarize_at(c("RepubFlip","DemoFlip","LineVal"), sum, na.rm  = TRUE)

county_st1 <- subset(county_st, RepubFlip >= 1)
county_st1$RFlipPct <- (county_st1$RepubFlip/county_st1$LineVal) * 100


county_st2 <- subset(county_st, DemoFlip >= 1)
county_st2$DFlipPct <- (county_st2$DemoFlip/county_st2$LineVal) * 100



county_st1$ER_states <- factor(county_st1$ER_states, levels = 
                                   c(as.character(unique(county_st1$ER_states))))

county_st2$ER_states <- factor(county_st2$ER_states, levels = 
                                 c(as.character(unique(county_st2$ER_states))))

p <- plot_ly(data = county_st1,
             x = county_st1$ER_states, 
             y = county_st1$RFlipPct,
             type = "scatter",
             size = county_st1$RFlipPct,
             name = '% Flip By Republicans')%>%
  add_trace(data = county_st2,
            x = county_st2$ER_states, 
            y = county_st2$DFlipPct,
            type = "scatter",
            size = county_st2$RFlipPct,
            name = '% Flip By Democrats')%>%
  layout(yaxis = list(title = ' % Counties Flipped in 2016'))%>%
  layout(xaxis = list(title = 'States')); p




```

&nbsp;
&nbsp;

#### Inference:

Above 2 plots tells us that during 2016 presidential elections, some states saw a trend favouring Republican candidate. Interesting to see that almost all of these states are either from North and Northeast US.

Below analysis for states, where significant county flips happened will highlight some trends.
     
     

&nbsp;
&nbsp;
&nbsp;

### Flipped State Analysis 

Based on above plots, following states are taken into account for flip state analysis:

    - NY
    - NH
    - ME
    - WI
    - MI
    - IA
    - MN
    - OH
    

&nbsp;
&nbsp;

```{r code4, warning=FALSE, message=FALSE,echo=FALSE}


swing.states1 <- subset(ER1, ER1$ER_states == 'WI' |ER1$ER_states == 'MI'| 
                             ER1$ER_states =='MN' |ER1$ER_states == 'OH'|
                             ER1$ER_states == 'IA'|ER1$ER_states == 'ME'|
                             ER1$ER_states == 'NH'|ER1$ER_states == 'NY')


##Identifying Counties flipped by Trump in 2016 election for WI state

county_flipRSW = subset(swing.states1,swing.states1$Winner2016 == 'Trump');
county_flipRSW = subset(county_flipRSW,county_flipRSW$Winner2012 == 'Obama')
county_Demo_WinnerSW = subset(swing.states1,swing.states1$Winner2016 == 'Clinton')
county_Rep_WinnerSW = subset(swing.states1,swing.states1$Winner2016 == 'Trump')



swing.states1$ER_states <- factor(swing.states1$ER_states, levels = 
                                   c(as.character(unique(swing.states1$ER_states))))

county_flipRSW$ER_states <- factor(county_flipRSW$ER_states, levels = 
                                    c(as.character(unique(county_flipRSW$ER_states))))

county_Demo_WinnerSW$ER_states <- factor(county_Demo_WinnerSW$ER_states, levels = 
                                    c(as.character(unique(county_Demo_WinnerSW$ER_states))))

county_Rep_WinnerSW$ER_states <- factor(county_Rep_WinnerSW$ER_states, levels = 
                                    c(as.character(unique(county_Rep_WinnerSW$ER_states))))




p3 <- plot_ly(county_flipRSW, 
             x = county_flipRSW$ER_states, 
             type = "histogram",
             name = 'flip')%>%
     add_trace(county_Demo_WinnerSW, x = county_Demo_WinnerSW$ER_states, name = "Clinton"  ) %>%
     add_trace(county_Rep_WinnerSW, x = county_Rep_WinnerSW$ER_states, name = "Trump"  ) %>%
     add_trace(swing.states1, x = swing.states1$ER_states, name  = 'Total Counties') %>%
     layout(yaxis = list(title = '# counties in Swing States'))%>%
     layout(xaxis = list(title = 'Swing States in 2016')); p3


cat("Total Number of Counties Flipped by Trump in Selected States      :", nrow(county_flipRSW))

```

&nbsp;
&nbsp;

#### Inference: 

It seems clear enough from plot that for few states majority counties were won by Trump, as a result of flip. Although, counties and electoral college is different but this plot is good to compare of flips and impact to Democratic party in swing states.


&nbsp;
&nbsp;
&nbsp;


#### Ethnic Population Split by Flipped States
 
  
&nbsp;
&nbsp;
    
```{r code5, warning=FALSE, message=FALSE,echo=FALSE}



ethn <- data.frame(States = swing.states1$ER_states, 
                   White.Pop = swing.states1$ER_Ethn_White,
                   Black.Pop = swing.states1$ER_Ethn_Black,
                   AmInd.Pop = swing.states1$ER_Amer_Ind,
                   Asian.Pop = swing.states1$ER_Asian,
                   haw.Pop = swing.states1$ER_Hawai,
                   Misc.Pop = swing.states1$ER_Misc,
                   Hisp.Pop = swing.states1$ER_Hispanic)

options(dplyr.print_max = 25)
options(pillar.sigfig=5)


ethn.sum <- ethn %>%
  group_by(States) %>%
  summarize_at(c("White.Pop","Black.Pop","AmInd.Pop",
                 "Asian.Pop", "haw.Pop", "Misc.Pop",
                 "Hisp.Pop"), sum, na.rm  = TRUE)

library(reshape2)
ethn.sum.long<-melt(ethn.sum,id.vars="States")


p1 <- plot_ly(subset(ethn.sum.long,ethn.sum.long$States == 'ME'), values = ~value,
             labels = ~variable, type = 'pie',hole = 0.40,
             domain = list(x = c(0, 0.5), y = c(0, 1))) %>%
      add_trace(data = subset(ethn.sum.long,ethn.sum.long$States == 'NH'), values = ~value,
             labels = ~variable, type = 'pie',hole = 0.40,
            domain = list(x = c(0.5, 1), y = c(0, 1)))%>%
      layout(title = 'Maine State                            New Hampshire State ');p1

p2 <- plot_ly(subset(ethn.sum.long,ethn.sum.long$States == 'NY'), values = ~value,
            labels = ~variable, type = 'pie',hole = 0.40,
            domain = list(x = c(0, 0.5), y = c(0, 1))) %>%
      add_trace(data = subset(ethn.sum.long,ethn.sum.long$States == 'WI'), values = ~value,
            labels = ~variable, type = 'pie',hole = 0.40,
            domain = list(x = c(0.5, 1), y = c(0, 1))) %>%
      layout(title = 'New York State                         Wisconsin State '); p2

p3 <- plot_ly(subset(ethn.sum.long,ethn.sum.long$States == 'MI'), values = ~value,
              labels = ~variable, type = 'pie',hole = 0.40,
              domain = list(x = c(0, 0.5), y = c(0, 1))) %>%
      add_trace(data = subset(ethn.sum.long,ethn.sum.long$States == 'MN'), values = ~value,
            labels = ~variable, type = 'pie',hole = 0.40,
            domain = list(x = c(0.5, 1), y = c(0, 1)))%>%
     layout(title = 'Michigan State                              Minnesota State '); p3

p4 <- plot_ly(subset(ethn.sum.long,ethn.sum.long$States == 'IA'), values = ~value,
              labels = ~variable, type = 'pie',hole = 0.40,
              domain = list(x = c(0, 0.5), y = c(0, 1))) %>%
      add_trace(data = subset(ethn.sum.long,ethn.sum.long$States == 'OH'), values = ~value,
            labels = ~variable, type = 'pie',hole = 0.40,
            domain = list(x = c(0.5, 1), y = c(0, 1)))%>%
      layout(title = 'Iowa State                              Ohio State '); p4
  

```

&nbsp;
&nbsp;


&nbsp;
&nbsp;
&nbsp;



#### College Population in Flipped States

&nbsp;
&nbsp;
  
```{r code7, warning=FALSE, message=FALSE,echo=FALSE}

## College Population in Swing States

p <- plot_ly(swing.states1, 
             x = subset((swing.states1$ER_College_Pop/swing.states1$ER_Pop14) *100, 
                         swing.states1$ER_states == 'WI'),
             type = "box", name = 'Wisconsin State') %>%
   add_trace(x = subset((swing.states1$ER_College_Pop/swing.states1$ER_Pop14) *100, 
                        swing.states1$ER_states == 'MN'),
             type = "box", name = 'Minnesota State') %>%
   add_trace(x = subset((swing.states1$ER_College_Pop/swing.states1$ER_Pop14) *100, 
                        swing.states1$ER_states == 'MI'),
             type = "box", name = 'Michigan State') %>%
   add_trace(x = subset((swing.states1$ER_College_Pop/swing.states1$ER_Pop14) *100, 
                        swing.states1$ER_states == 'IA'),
             type = "box", name = 'Iowa State') %>%
   add_trace(x = subset((swing.states1$ER_College_Pop/swing.states1$ER_Pop14) *100, 
                        swing.states1$ER_states == 'OH'),
             type = "box", name = 'Ohio State') %>%
   add_trace(x = subset((swing.states1$ER_College_Pop/swing.states1$ER_Pop14) *100, 
                        swing.states1$ER_states == 'NY'),
             type = "box", name = 'New York State') %>%
   add_trace(x = subset((swing.states1$ER_College_Pop/swing.states1$ER_Pop14) *100, 
                       swing.states1$ER_states == 'NH'),
            type = "box", name = 'New Hampshire State') %>%
   add_trace(x = subset((swing.states1$ER_College_Pop/swing.states1$ER_Pop14) *100, 
                       swing.states1$ER_states == 'ME'),
            type = "box", name = 'Maine State') %>%
   layout(xaxis = list(title = 'College Population % in Swing States')); p

```
&nbsp;
&nbsp;

#### Inference: 

From above plot, it is quite clear that Northern states have comparativley low % of college population.



&nbsp;
&nbsp;
&nbsp;

### State vs Voter Turnaround
&nbsp;
&nbsp;


#### Assumptions:

Age Eligible Voting population is considered for below analysis. In reality actual voting population is the registered voter population. 

Voting Age Populaiton consists of Registered voters + Nonvoters + Noncitizens

```{r code8, warning=FALSE, message=FALSE,echo=FALSE}


options(dplyr.print_max = 25)
options(pillar.sigfig=5)
a2 <- ER1 %>%
  group_by(ER_states) %>%
  summarize(ER_Vot_Turnarnd = mean(ER_Vot_Turnarnd))

a3 <- ER1 %>%
  ##group_by(ER_states) %>%
  summarize(ER_Vot_Turnarnd = mean(ER_Vot_Turnarnd))

p2 <- plot_ly(data = a2,
            x = a2$ER_states,
            y = a2$ER_Vot_Turnarnd,
            type = "scatter",
            mode = 'lines+markers',
        ##    color = ('rgb(0, 0, 255)'),
            name = 'State Voter Turnaround') %>%
  add_trace(data = a3,
            y = a3$ER_Vot_Turnarnd,
            type = "scatter",
            mode = 'lines',
            ##    color = ('rgb(200, 100, 50)'),
            name = 'US Voter Turnaround'); p2

```
&nbsp;
&nbsp;


#### Inference: 

Observing the graph, newyork state saw low voter turnout but other states, where we saw significant flip in counties towards republicans has significantly higher voter turnout compared to mean US voter turnout. Some of these flip states have higher % of white population, which tells us that white voter turnout is higher in these flipped states.   

   
&nbsp;
&nbsp;
&nbsp;

#### President Candidate Vote % Across all US Counties
&nbsp;
&nbsp;
```{r code10, warning=FALSE, message=FALSE,echo=FALSE}


p2 <- plot_ly(ER1, x = ER1$ER_Clinton, type = "box", name = 'Clinton') %>%
        add_trace(x = ER1$ER_Trump, name  = 'Trump') %>%
        add_trace(x = ER1$ER_Obama, name  = 'Obama') %>%
        add_trace(x = ER1$ER_Romney, name  = 'Romney') %>%
  layout(xaxis = list(title = 'Pres. Cand. Vote %')); p2

```

&nbsp;
&nbsp;



&nbsp;
&nbsp;
&nbsp;
  
  


## Bivariate and Multivariate Analysis

&nbsp;
&nbsp;

### Identity Politics 

These days politics is being defined by characteristics of the voting population. Below analysis will highlight some key trends and establish correlation between county level vote % and important variables.

Both 2012 and 2016 election results are being considered.


&nbsp;
&nbsp;

#### College Population Correlation with Vote casted
&nbsp;
&nbsp;
```{r code21, warning=FALSE, message=FALSE,echo=FALSE}

### Correlation - 2016

f <- list(
family = "Arial",
size = 15,
color = 'black')

p1 <- plot_ly(ER,
              x = ~Clinton,
              y = ~Edu_batchelors,
              type = "scatter",
              name = 'Clinton') %>%
  add_trace(x = ~Trump, name = 'Trump') %>%
  layout(title = '2012-2016 Candidates Vs College Educated Pop%', titlefont = f,
         yaxis = list(title = 'County College Pop. %',zeroline = FALSE, titlefont = f),
         xaxis = list(title = 'Vote Share %',zeroline = FALSE, titlefont = f));p1

### Correlation - 2012

p2 <- plot_ly(ER,
              x = ~Obama,
              y = ~Edu_batchelors,
              type = "scatter",
              name = 'Obama') %>%
  add_trace(x = ~Romney, name = 'Romney') %>%
  layout(title = '2012-2016 Candidates Vs College Educated Pop%', titlefont = f,
         yaxis = list(title = 'County College Pop. %',zeroline = FALSE, titlefont = f),
         xaxis = list(title = 'Vote Share %',zeroline = FALSE, titlefont = f));p2

##p <- subplot(p1, p2); p

```

&nbsp;
&nbsp;

#### Inference: 

Over last 2 elections, we can clearly see the widening split among college educated and non college educated population.

   

&nbsp;
&nbsp;
&nbsp;

#### Household Income Correlation with Vote casted
&nbsp;
&nbsp;
```{r code22, warning=FALSE, message=FALSE,echo=FALSE}


p4 <- plot_ly(ER,
              x = ~Clinton,
              y = ~INC110213,
              type = "scatter",
              name = 'Clinton') %>%
  add_trace(x = ~Trump, name = 'Trump') %>%
  layout(title = '2012-2016 Candidates Vs Median Household Income', titlefont = f,
         yaxis = list(title = 'Median Household Income',zeroline = FALSE, titlefont = f),
         xaxis = list(title = 'Vote Share %',zeroline = FALSE, titlefont = f));p4

### Correlation - 2012

p5 <- plot_ly(ER,
              x = ~Obama,
              y = ~INC110213,
              type = "scatter",
              name = 'Obama') %>%
  add_trace(x = ~Romney, name = 'Romney') %>%
  layout(title = '2012-2016 Candidates Vs Median Household Income', titlefont = f,
         yaxis = list(title = 'Median Household Income',zeroline = FALSE, titlefont = f),
         xaxis = list(title = 'Vote Share %',zeroline = FALSE, titlefont = f));p5

##p6 <- subplot(p4, p5); p6



```

&nbsp;
&nbsp;

#### Inference: 

Over last 2 elections, we can clearly see the widening split based on Household income. 


&nbsp;
&nbsp;
&nbsp;



#### Gender (Female Population) Correlation with Vote casted
&nbsp;
&nbsp;

```{r code23, warning=FALSE, message=FALSE,echo=FALSE}
p7 <- plot_ly(ER,
              x = ~Clinton,
              y = ~SEX255214,
              type = "scatter",
              name = 'Clinton') %>%
  add_trace(x = ~Trump, name = 'Trump') %>%
  layout(title = '2012-2016 Candidates Vs Female Voters', titlefont = f,
         yaxis = list(title = 'Female % Population',zeroline = FALSE, titlefont = f),
         xaxis = list(title = 'Vote Share %',zeroline = FALSE, titlefont = f));p7

### Correlation - 2012

p8 <- plot_ly(ER,
              x = ~Obama,
              y = ~SEX255214,
              type = "scatter",
              name = 'Obama') %>%
  add_trace(x = ~Romney, name = 'Romney') %>%
  layout(title = '2012-2016 Candidates Vs Female Voters', titlefont = f,
         yaxis = list(title = 'Female % Population',zeroline = FALSE, titlefont = f),
         xaxis = list(title = 'Vote Share %',zeroline = FALSE, titlefont = f));p8

##p9 <- subplot(p7, p8); p9


```

&nbsp;
&nbsp;


#### Inference: 

Over last 2 elections, there seems no significant change.

 
  
    
 &nbsp;
 &nbsp;
 &nbsp;

#### Ethnicity Correlation with Vote casted

&nbsp;
&nbsp;

```{r code24, warning=FALSE, message=FALSE,echo=FALSE}
### Correlation - 2016

p10 <- plot_ly(ER,
              x = ~Clinton,
              y = ~White,
              type = "scatter",
              name = 'Clinton-White') %>%
  add_trace(x = ~Trump, name = 'Trump-White') %>%
  layout(title = 'White Population Split', titlefont = f,
         yaxis = list(title = '% White Pop.',zeroline = FALSE, titlefont = f),
         xaxis = list(title = 'Vote Share %',zeroline = FALSE, titlefont = f));p10

### Correlation - 2012

p11 <- plot_ly(ER,
              x = ~Clinton,
              y = ~Black,
              type = "scatter",
              name = 'Clinton-Black') %>%
  add_trace(x = ~Trump, name = 'Trump-Black') %>%
  layout(title = 'Black population Split', titlefont = f,
         yaxis = list(title = '% Black Pop.',zeroline = FALSE, titlefont = f),
         xaxis = list(title = 'Vote Share %',zeroline = FALSE, titlefont = f));p11


```

 &nbsp;
 &nbsp;

#### Inference: 


For 2016 election, we can clearly see the wide split between ethnicity and political party.
 
   
   

&nbsp;
&nbsp;
&nbsp;


### Examining the Age Distribution in each state

&nbsp;
&nbsp;

```{r code9, warning=FALSE, message=FALSE,echo=FALSE}

#Calculating Mean Age category % of Every State

options(dplyr.print_max = 25)
options(pillar.sigfig=5)
Age.Dist <- ER %>%
  group_by(state_abbr) %>%
  summarize_at(c("AGE135214","AGE295214","age65plus"),
                 mean, na.rm  = TRUE)

#Calculating Mean Age Category % of USA

Age.Dist1 <- ER %>%
 ## group_by(state_abbr) %>%
  summarize_at(c("AGE135214","AGE295214","age65plus"),
                 mean, na.rm  = TRUE)

## Plotting Distribution of State Mean Age Categories

p19 <- plot_ly(Age.Dist) %>% 
    add_trace(x = ~state_abbr,
            y = ~AGE135214,
            type = "scatter",
            mode = 'lines+markers',
            color = ('rgb(0, 0, 255)'),
            name = 'Age Under 5 %')%>%
    add_trace(x = ~state_abbr,
            y = Age.Dist1$AGE135214,
            type = "scatter",
            mode = "lines",
            color = ('rgb(221,160,221)'),
            hoverinfo = "none",
            name = 'Age Under 5 % - USA')%>%
    add_trace(x = ~state_abbr,
            y = ~AGE295214,
            type = "scatter",
            mode = 'lines+markers',
            color = ('rgb(0,100,0)'),
            name = 'Age 5 to 18 %')%>%
    add_trace(x = ~state_abbr,
            y = Age.Dist1$AGE295214,
            type = "scatter",
            mode = "lines",
            color = ('rgb(221,160,221)'),
            hoverinfo = "none",
            name = 'Age 5 to 18 % - USA')%>%
    add_trace(x = ~state_abbr,
            y = (100 - (Age.Dist$AGE135214 + Age.Dist$AGE295214 + Age.Dist$age65plus)),
            type = "scatter",
            mode = 'lines+markers',
            color = ('rgb(255,0,0)'),
            name = 'Age 18 to 65 %')%>%
    add_trace(x = ~state_abbr,
            y = (100 - (Age.Dist1$AGE135214 + Age.Dist1$AGE295214 + Age.Dist1$age65plus)),
            type = "scatter",
            mode = "lines",
            color = ('rgb(221,160,221)'),
            hoverinfo = "none",
            name = 'Age 18 to 65 % - USA')%>%
    add_trace(x = ~state_abbr,
            y = ~age65plus,
            type = "scatter",
            mode = 'lines+markers',
            color = ('rgb(148,0,211)'),
            name = 'Age 65+ %')%>%
    add_trace(x = ~state_abbr,
            y = Age.Dist1$age65plus,
            type = "scatter",
            mode = "lines",
            color = ('rgb(221,160,221)'),
            hoverinfo = "none",
            name = 'Age 65+ % - USA')%>%
    layout(title = 'State Average Age Distribution',
            xaxis = list(title = 'State'),
            yaxis = list(title = 'Population % - Age')); p19

```

 &nbsp;
 &nbsp;
  

&nbsp;
&nbsp;
&nbsp;

#### Analyzing Point Differenential between 2016 and 2012 presidential elections.

&nbsp;

Observing county-level results (in terms of % margin) distribution:

&nbsp;

```{r code12, warning=FALSE, message=FALSE,echo=FALSE}

##summary(ER1$ER_diff2016)
x1 <- density(ER1$ER_diff2016)
x2 <- density(ER1$ER_diff2012)

p2 <- plot_ly(ER1, x = ER1$ER_diff2016, 
              type = "histogram",
              name = '2016') %>%
  add_trace(x = ER1$ER_diff2012,
            type = "histogram",
            name = '2012') %>%
  add_trace(x = x1$x, y = x1$y, type = "scatter", 
            mode = "lines", fill = "blue", yaxis = "y2", name = "Density") %>% 
  add_trace(x = x2$x, y = x2$y, type = "scatter", 
            mode = "lines", fill = "orange", yaxis = "y2", name = "Density") %>% 
   layout(  xaxis = list(title = 'Point Differential: Democrats - Republicans'),
            yaxis = list(title = 'No. Of Counties')) %>% 
   layout(yaxis2 = list(overlaying = "y", side = "right")); p2


```

&nbsp;
&nbsp;

#### Inference: 

We can see that the histogram moved to the left, in Trump's favor. 

Republican Candidates won the vast majority of counties, and lost the popular vote. This is due to the fact that counties can differ in population size,and the biggest counties - the cities - usually go with the democrats.


 
 &nbsp;
 &nbsp;
 &nbsp;
 &nbsp;

## Central Limit Theorem Application

The central limit theorem is perhaps the most fundamental result in all of statistical theory, which states that given a sufficiently large sample size from a population with a finite level of variance, the mean of all samples from the same population will be approximately equal to the mean of the population.
It allows us to understand the behavior across repeated sampling and thereby conclude if a result from a given sample can be declared to be statistically important. 
Specifically, as the sample sizes get larger, the distribution of means calculated from repeated sampling will approach normality.  
Three different components of the central limit theorem are successive sampling from a population, increase in sample size and population distribution.



&nbsp;
&nbsp;

### Getting Distribution of Income across US States
&nbsp;

```{r code14, warning=FALSE, message=FALSE,echo=FALSE}
##swing.states.PA <- subset(swing.states1, swing.states1$ER_states == 'PA') 


p1 <- plot_ly(ER1, x = ER1$ER_House_income, 
              histnorm = "probability", 
              type = "histogram",
              name = 'Income') %>%
      layout(xaxis = list(title = 'Median Household Income'),
             yaxis = list(title = 'Density'));p1

m <- mean(ER1$ER_House_income)
std <-sd(ER1$ER_House_income)

##m <- mean(swing.states1$ER_House_income);m
##std <-sd(swing.states1$ER_House_income);std
```
&nbsp;
&nbsp;

&nbsp;
&nbsp;


##### Drawing 100 Samples for various sample sizes of 10,20,30,40
&nbsp;
```{r code15, warning=FALSE, message=FALSE,echo=FALSE}

set.seed(150)
par(mfrow = c(2,2))
samples <- 100
##sample.size <- 35

xbar <- numeric(samples)


for (sample.size in c(10,20,30,40)) {
for (i in 1: samples) {
xbar[i] <- mean(sample(ER1$ER_House_income,sample.size))
}


m1 = mean(xbar)
std1 <- sd(xbar)

hist(xbar,
breaks = 20,prob= TRUE,
main = paste("Sample Size =", sample.size),
xlab = "House Hold Income",col="Green")
curve(dnorm(x, mean=m1, sd=std1), 
col="red", lwd=2, add=TRUE, yaxt="n")


}


for (sample.size in c(10,20,30,40)) {
for (i in 1: samples) {
xbar[i] <- mean(sample(ER1$ER_House_income,sample.size))
}
  
cat("Sample Size = ", sample.size, " Mean = ", mean(xbar),
      " SD = ", sd(xbar), "\n")

}

cat("Sample Size = ", "OD", " Mean = ", m,
    " SD = ", std, "\n")


```
&nbsp;
&nbsp;


&nbsp;
&nbsp;
&nbsp;

## Sampling Techniques
&nbsp;

Sometimes, it is impractical to infer from an entire population. There are 157 million registered voters in the United States, which is quire a big umber. That is why appropriate samples are selected based on most appropriate sampling method, that can closely represents the whole population, with some margin of error. 


For this project, below 3 common sampling methods will be analyzed. Calculated field of 'Voter Turnaround Percentage' across US counties is considered for analysis.

Total Estimated Voting Eligible Population in US: 245.5 million
Total Estimated Registered Voters in US: 157.5 million


Source of above Information: 'https://www.census.gov/content/dam/Census/library/publications/2018/demo/P20-582.pdf'


&nbsp;
&nbsp;

### 1. Simple Random Sampling without Replacement
&nbsp;

```{r code40, warning=FALSE, message=FALSE,echo=FALSE}

set.seed(125)

S <- srswr(50,nrow(ER1))

##S[S != 0]
rows <- (1:nrow(ER1))[S !=0]

srswr.sample.ER1 <- ER1[rows,]

##Show the frequencies for each region (REG). 

a <- table(srswr.sample.ER1$ER_states)


###

##  Summarizing Voter Turnaround % at County Level for Simple Random Sampling Data

options(dplyr.print_max = 25)
options(pillar.sigfig=5)
a3 <- srswr.sample.ER1 %>%
  group_by(ER_states) %>%
  summarize(ER_Vot_Turnarnd = mean(ER_Vot_Turnarnd))


##  Summarizing Voter Turnaround % at County Level for Original Data

options(dplyr.print_max = 25)
options(pillar.sigfig=5)
a2 <- ER1 %>%
  group_by(ER_states) %>%
  summarize(ER_Vot_Turnarnd = mean(ER_Vot_Turnarnd))


## Merging Original Sumamrized Data with Simple Random Sampling Sumamrized Data

a4 <- merge(a2,a3,by="ER_states")


##Comparing Original Data with Simple Random Sampling of sample size 50

p30 <- plot_ly(a4) %>% 
      add_trace(x = a4$ER_states,
                y = a4$ER_Vot_Turnarnd.x,
                type = "scatter",
                mode = 'lines',
                name = 'Original Data') %>% 
      add_trace(x = a4$ER_states,
                y = a4$ER_Vot_Turnarnd.y,
                type = "scatter",
                mode = 'lines',
                name = 'Simple Random Sampling of 50');p30


```
&nbsp;
&nbsp;
 

  
   

&nbsp;
&nbsp;


### 2. Systematic Sampling
&nbsp;

```{r code41, warning=FALSE, message=FALSE,echo=FALSE}

set.seed(125)

N <- nrow(ER1)
n <- 50

k <- ceiling(N/n)

r <- sample(k,1)

# select every kth item

s1 <- seq(r, by = k, length = n)

sys.sample.ER1 <- ER1[s1, ]
##head(sys.sample.ER1)

for(i in 1:n) {
  if (s1[i] > N)
    s1[i] <- N
}

sys.sample.ER1 <- ER1[s1, ]

##  Summarizing Voter Turnaround % at County Level for Simple Random Sampling Data

options(dplyr.print_max = 25)
options(pillar.sigfig=5)
a5 <- sys.sample.ER1 %>%
  group_by(ER_states) %>%
  summarize(ER_Vot_Turnarnd = mean(ER_Vot_Turnarnd))


## Merging Original Sumamrized Data with Simple Random Sampling Sumamrized Data

a6 <- merge(a2,a5,by="ER_states")

##Comparing Original Data with Simple Random Sampling of sample size 50

p31 <- plot_ly(a6) %>% 
  add_trace(x = a6$ER_states,
            y = a6$ER_Vot_Turnarnd.x,
            type = "scatter",
            mode = 'lines',
            name = 'Original Data') %>% 
  add_trace(x = a6$ER_states,
            y = a6$ER_Vot_Turnarnd.y,
            type = "scatter",
            mode = 'lines',
            name = 'Systematic Sampling of 50');p31

```

&nbsp;
&nbsp;
 
 &nbsp;
 &nbsp;
 &nbsp;

### 3. Stratified sampling 
&nbsp;


Stratified Sampling method refers to dividing the population into separate groups, called strata. Then, a probability sample is drawn from each group. This method is useful when we donot want to ignore sample data from certain group.

Using Stratified Sampling, it may be possible to reduce the sample size required to achieve a given precision or to increase the precision with the same sample size. 

For this project, I divided the entire populaiton into below 6 categories based on voter turnout %:

     >  0% - 40%
     >  40% - 50%
     >  50% - 60%
     >  60% - 70%
     >  70% - 80%
     >  80% and Above

This will make sure that there will be atleast 1 sample picked up from each category, when sample size is 50.

&nbsp;
&nbsp;
```{r code42, warning=FALSE, message=FALSE,echo=FALSE}
set.seed(125)
#

## Setting Categories for Voter Turaround % based on sample size of 50, so that atleast 1
## sample get picked up from Stratified sampling

ER1$VoterTurn_Cat <- 
     ifelse(ER1$ER_Vot_Turnarnd < 40, "0% - 40%",
      ifelse(ER1$ER_Vot_Turnarnd >= 40 & ER1$ER_Vot_Turnarnd < 50, "40% - 50%",
       ifelse(ER1$ER_Vot_Turnarnd >= 50 & ER1$ER_Vot_Turnarnd < 60, "50% - 60%",
        ifelse(ER1$ER_Vot_Turnarnd >= 60 & ER1$ER_Vot_Turnarnd < 70, "60% - 70%",
         ifelse(ER1$ER_Vot_Turnarnd >= 70 & ER1$ER_Vot_Turnarnd < 80, "70% - 80%",                         ifelse(ER1$ER_Vot_Turnarnd >= 80 , "80% and Above",""))))))     

##order.index                                                                            
order.index <- order(ER1$VoterTurn_Cat)

##Arranging Data as per Index
data1 <- ER1[order.index, ]

##Getting Frequencies for each Category
freq <- table(data1$VoterTurn_Cat)

##Some States have more than 100% turnaround. Voter Fraud can be ruled out here. I guess County stat are from 2014 abd 2016 election happened in 2016. SO some under 18 might be eligible for votingv in 2016.

##Getting Proportions based on sample size 50
sizes <- (50 * freq / sum(freq))


```
```{r 47, warning=FALSE, message=FALSE, echo=FALSE, results="hide"}
st <- strata(data1, stratanames = c("VoterTurn_Cat"),
            size = round(sizes),
            method = "srswor",description = TRUE)

stf.sample.ER1<- getdata(data1, st)
##head(stf.sample.ER1)

## Summarizing Actual Voter Turnaround for sample data group by states

options(dplyr.print_max = 25)
options(pillar.sigfig=5)
a7 <- stf.sample.ER1 %>%
  group_by(ER_states) %>%
  summarize(ER_Vot_Turnarnd = mean(ER_Vot_Turnarnd))

## Merginf Stratified sample and Original sumamrized data for Voter Turnaround %

a8 <- merge(a2,a7,by="ER_states")
```

```{r code48, warning=FALSE, message=FALSE,echo=FALSE}
p32 <- plot_ly(a8) %>% 
  add_trace(x = a8$ER_states,
            y = a8$ER_Vot_Turnarnd.x,
            type = "scatter",
            mode = 'lines',
            name = 'Original Data') %>% 
  add_trace(x = a8$ER_states,
            y = a8$ER_Vot_Turnarnd.y,
            type = "scatter",
            mode = 'lines',
            name = 'Stratified Sampling of 50');p32

```


&nbsp;
&nbsp;
&nbsp;


#### Sampling Techniques Conclusion 

Surpisingly, Simple Random sampling without replacement curve follows the curve of original data very closely.

Statified sampling is the closest match as compared to original distribution of voter turnout % but even that sampling method is seeing some significant deviations for states which were selected in sampling.

Systematic sampling do have some very close match for some states but many small states with less number of counties can be completely ignored by this sampling method.



If above samples are used for data analysis, some of the key facts can be overlooked completely. As the dataset is not itself very huge, getting smaller samples from dataset will not be good for any inferential analysis.

&nbsp;
&nbsp;
&nbsp;


## Conclusion

In this project analysis, univriate and bivariate analysis were performed on various variables to understand the trends from 2012 and 2016 Presidential results. Some key co-relations and trends were observed during county level bivriate analysis. Analysis was performed on states where significant counties were flipped to understand the characteristics of populaiton at county level. Numerical Distribution on differential vote % highlights that there has been positive gain towards republican party.







